---
title: "Further details ggrmsMD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Further_details_ggrmsMD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup, warning=FALSE, message=FALSE}
library(rms)
library(rmsMD)
```

# Introduction and setup

This vignette goes into further details for the `ggrmsMD` function. For further package information and basic use please refer to the vignette **Standard_workflow_with_restricted_cubic_splines**.

For these vignettes we will use a simulated dataset to predict the impact of age, BMI, Sex and Smoking status on outcome after surgery. The models are for illustration purposes only.

```{r}

# Load in the simulated data
data <- simulated_rmsMD_data()

# Set the datadist which is required for rms modelling (these two lines are standard)
dd <- datadist(data)    
options(datadist='dd') 

```

```{r}

# Fit an OLS model including a restricted cubic spline for Age and BMI (with 4 knots)
fit_lrm <- lrm(majorcomplication ~ rcs(age,4) + rcs(bmi,4) + sex + smoking, data = data)

```

## Plotting with ggrmsMD

Now that the model and overall p values have been assessed, the `ggrmsMD` function from **rmsMD** can be used to assess the relationship between variables modelled with restricted cubic splines, and the outcome.

As a minimum, the model fit and data should be passed into the function. `ggrmsMD` will then generate plots for all variables which were modelled with restricted cubic splines. By default: for linear regression models predicted outcome is plotted, for logistic regression OR is plotted and for Cox regression HR is plotted. All of these plots are adjusted for all other variables in the model.

All of the outputted plots are ggplots, and therefore can be further adapted using that framework.

Here is the most basic use case with the logistic regression model for post-operative complications above:

```{r}
# most basic output
ggrmsMD(fit_lrm, data)

```

`ggrmsMD` has several additional options to improve these plot outputs for publications. Some of these options are shown below. For the examples below, this is done iteratively for demonstration purposes.

Arguments which alter the y axis are applied to all plots. This is standardised as all plots have originated from the same regression model:

```{r}

ggrmsMD(fit_lrm, data,
        ylab = "Complications (adjusted OR)", # custom y axis label 
        ylim = c(0,3) # set y axis limits so variables can be compared more easily
        )

```

Shading can be applied to these plots to make it clear which side of the no-effect line represents inferior/superior clinical outcome. This is achieved by setting `shade_inferior` to "higher" or "lower":

```{r}
ggrmsMD(fit_lrm, data,
        ylab = "Complications (adjusted OR)", 
        ylim = c(0,3), 
        shade_inferior = "higher" # define that higher OR mean inferior outcome to add red shading
        )
```

It may be preferable to have the y-axis on a log-scale rather than a linear scale. For logistic regression models like this one, predicted probability can be plotted instead of adjusted OR. For example:

```{r}

# to have the y axis on a log-scale:

ggrmsMD(fit_lrm, data,
        ylab = "Complications (adjusted OR)", 
        ylim = c(0.25,4), 
        shade_inferior = "higher",
        log_y = TRUE, # have the y-axis on a log scale
        log_y_breaks = c(0.25, 0.5 , 1, 2, 4) # optionally set the y-axis breaks
        )

# to plot predicted probability rather than adjusted OR. This is adjusted to mode/mean of the other variables

ggrmsMD(fit_lrm, data,
        ylim = c(0,0.4), 
        lrm_prob = TRUE # set this to plot predicted probability
        )

```

Changes applied to the x-axis and titles all apply to each plot individually (rather than being common accross all plots like the y-axis changes). This is achieved using lists. For example to alter the x-axis labels for both variables, and x-axis limits for only age:

```{r}

xlabels <- list ("age" = "Age (years)",
                 "bmi" = "Body Mass Index")

xlimits <- list("age" = c(35,65))

ggrmsMD(fit_lrm, data,
        ylab = "Complications (adjusted OR)", 
        ylim = c(0,3), 
        shade_inferior = "higher",
        xlabs = xlabels,
        xlims = xlimits
        )

```

The same approach is used if titles are required:

```{r}

xlabels <- list ("age" = "Age (years)",
                 "bmi" = "Body Mass Index")

titles <- list ("age" = "Impact of Age on Complications",
                 "bmi" = "Impact of BMI on Complications")

ggrmsMD(fit_lrm, data,
        ylab = "Complications (adjusted OR)", 
        ylim = c(0,3), 
        shade_inferior = "higher",
        xlabs = xlabels,
        titles = titles
        )

```

The x-axis can also be displayed on a log-scale, rather than a linear scale. This is done using `log_x_vars` to specify which variables to have log-scaled x-axes:

```{r}

ggrmsMD(fit_lrm, data,
        ylab = "Complications (adjusted OR)", 
        ylim = c(0,3), 
        shade_inferior = "higher",
        xlabs = xlabels,
        titles = titles,
        log_x_vars = c("age", "bmi")
        )

```

`combined` is set to `TRUE` by default to output a single combined plot, as shown above. `combined = FALSE` can be used to output a list of plots. If a single plot is required, this can be achieved with the `var` argument:

```{r}

ggrmsMD(fit_lrm, data,
        ylab = "Complications (adjusted OR)", 
        ylim = c(0,3), 
        shade_inferior = "higher",
        var = "age"
        )

```

For linear regression models the ouput is predicted outcome. For example, here for length of stay:

```{r}

fit_ols <- ols(lengthstay ~ rcs(age,4) + rcs(bmi,4) + sex + smoking, data = data)

ggrmsMD(fit_ols, data,
        ylim = c(20,50),
        ylab = "Predicted length of stay"
        )

```


