---
title: "modelsummary_rms vignettes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{modelsummary_rms-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

The **modelsummary_rms** function is designed to process output from models fitted using the **rms** package and generate a summarised dataframe of the results. 
This vignette will guide you through the basic usage of the function and then move on to more advanced examples.
The goal is to produce publication-ready summaries of the models.

## Installation and Setup

Make sure you have the required packages installed. 
You can install the package (if available on CRAN or GitHub) along with its dependencies:

Note that you may also need to install other packages (such as flextable and officer) if you plan to convert the output dataframe into publication-ready tables.
```{r}
# Install the package if you haven't already
# install.packages("rmsMD")
# or for GitHub:
# devtools::install_github("https://github.com/sjtingle/rmsMD")

library(rms)
library(rmsMD)
```

# Basic Usage

Here is a simple example using an ordinary least squares (OLS) model with the built-in survey dataset from the MASS package:
```{r}
# Using the built-in dataset from the MASS package
data("survey", package = "MASS")

# Fit an OLS model using the rms package
fit_ols <- ols(Wr.Hnd ~ Age + Exer + Sex, data = survey)

# Generate a model summary
summary_df <- modelsummary_rms(fit_ols)
knitr::kable(summary_df)
```
The output dataframe contains the estimated coefficients, their 95% confidence intervals, and the associated p-values.

## Customising the Output

By default, the function uses the following stylistic settings:

- **combine_ci = TRUE:** Combines the effect estimate and the 95% confidence interval into a single column.
- **round_dp_coef = 3:** Rounds the effect estimates to three decimal places.
- **round_dp_p = 3:** Rounds the p-values to three decimal places.

You can modify these defaults to adjust the appearance of the output.
```{r}
# Generate a model summary with custom styling options
summary_custom <- modelsummary_rms(fit_ols, 
                                   combine_ci = FALSE, 
                                   round_dp_coef = 2, 
                                   round_dp_p = 5)
knitr::kable(summary_custom)
```
## Full Model Output

By default, **modelsummary_rms** returns only the final formatted summary (i.e. `fullmodel = FALSE`). 
This output made to be concise and show the key results.

If all information is required, you can set `fullmodel = TRUE`. 
This option returns additional results.

```{r}
# Generate a model summary with custom styling options
summary_fullmodel <- modelsummary_rms(fit_ols, 
                                   combine_ci = FALSE, 
                                   round_dp_coef = 2, 
                                   round_dp_p = 5,
                                   fullmodel = TRUE)
knitr::kable(summary_fullmodel)
```

## Exponential Coefficients

Exponentiating the coefficients of certain models makes the interpretation more intuitive (e.g. as odds ratios in logistic regression and hazard ratios in Cox models). 
The **rmsMD** package is able to automatically set the appropriate value for `exp_coef` for RMS models such as `ols`, `lrm`, and `cph`. 

Below is an example using an RMS OLS model and an RMS logistic regression model that both work with the default settings for `exp_coef`:
```{r}
# Example using an RMS OLS model
data("survey", package = "MASS")
fit_ols <- ols(Wr.Hnd ~ Age + Exer + Sex, data = survey)
summary_ols <- modelsummary_rms(fit_ols)
knitr::kable(summary_ols)
```

```{r}
# Example using an RMS logistic regression model
# Note: For demonstration, we create a binary outcome using the survey dataset.
survey$BinaryOutcome <- ifelse(survey$Wr.Hnd > median(survey$Wr.Hnd, na.rm = TRUE), 1, 0)
fit_lrm <- lrm(BinaryOutcome ~ Age + Exer + Sex, data = survey)
summary_lrm <- modelsummary_rms(fit_lrm)
knitr::kable(summary_lrm)
```


# Non rms Models

The **rmsMD** package is also capable of working with non-RMS models, such as those fitted using base R functions like lm(). 
However, in these cases the package cannot automatically determine the appropriate value for exp_coef, so you must manually set it. 
For example, when using a linear model (where exponentiation is not required), you should explicitly set exp_coef = FALSE.
```{r}
# Fit a simple linear model using lm() from base R (non-RMS model)
fit_lm <- lm(Wr.Hnd ~ Age + Exer + Sex, data = survey)

# Generate a model summary for the non-RMS model by explicitly setting exp_coef = FALSE
summary_lm <- modelsummary_rms(fit_lm, 
                               exp_coef = FALSE)
knitr::kable(summary_lm)
```

# Restricted Cubic Splines

Restricted Cubic Splines (RCS) are a flexible modelling tool used to capture non-linear relationships between predictors and outcomes. 
In medicine, many continuous variables (e.g. age, blood pressure, or biomarker levels) do not have a strictly linear association with the outcome of interest. 
RCS analysis allows for this to be modelled non-linearly, making it a particularly useful tool in clinical research.

The **rmsMD** package is designed to report and summarise models that include RCS terms. 
It extracts the effect estimates and confidence intervals for the RCS components and also provides options to:

- **Display an overall p-value for the spline terms using the `rcs_overallp` option.**  
  When this option is set to `TRUE`, the function computes a single p-value that tests the overall significance of the spline terms. 
  This overall p-value provides insight into whether the relationship between the predictor and the dependent variable is significant and whether it deviates from a simple linear relationship. 
  In other words, a significant overall p-value suggests that the non-linear terms add meaningful information beyond a linear effect.

- **Hide the individual spline coefficients using the `hide_rcs_coef` option.**  
  Hiding the detailed spline coefficients can be advantageous because these individual parameters often lack straightforward clinical interpretation. 
  Instead, the focus is on the overall association captured by the spline. The overall p-value (when `rcs_overallp` is `TRUE`) summarises the significance of the non-linear relationship, allowing you to conclude whether the predictor’s effect is simply linear or if a non-linear relationship is present and significant. 
  This helps simplify the output by emphasizing the broader pattern rather than the intricate details of the spline transformation.

## Basic RCS Model without Hiding Coefficients

In this example, we fit an OLS model that includes an RCS term for Age. 
The default setting shows all coefficients including those for the spline terms.
```{r}
# Using the built-in dataset from the MASS package
data("survey", package = "MASS")

# Fit an OLS model including a restricted cubic spline for Age (with 4 knots)
fit_spline <- ols(Wr.Hnd ~ rcs(Age, 4) + Exer + Sex, data = survey)

# Generate a model summary without modifying RCS settings (default: rcs_overallp = FALSE, hide_rcs_coef = FALSE)
summary_spline <- modelsummary_rms(fit_spline)
knitr::kable(summary_spline)
```

## RCS Model with Overall p-value and Hiding Individual Coefficients

This is particularly useful when you want to focus on the significance of the non-linear relationship.
```{r}
# Fit an OLS model including a restricted cubic spline for Age (with 4 knots)
fit_spline_hide <- ols(Wr.Hnd ~ rcs(Age, 4) + Exer + Sex, data = survey)

# Generate a model summary with rcs_overallp set to TRUE and hide_rcs_coef set to TRUE
summary_spline_hide <- modelsummary_rms(fit_spline_hide, 
                                        rcs_overallp = TRUE, 
                                        hide_rcs_coef = TRUE)
knitr::kable(summary_spline_hide)
```

## RCS Model with Overall p-value and Individual Coefficients

```{r}
# Fit an OLS model including a restricted cubic spline for Age (with 4 knots)
fit_spline_hide <- ols(Wr.Hnd ~ rcs(Age, 4) + Exer + Sex, data = survey)

# Generate a model summary with rcs_overallp set to TRUE and hide_rcs_coef set to TRUE
summary_spline_hide <- modelsummary_rms(fit_spline_hide, 
                                        rcs_overallp = TRUE, 
                                        hide_rcs_coef = FALSE)
knitr::kable(summary_spline_hide)
```

# Models with Interactions

Interactions occur when the effect of one predictor on the dependent variable depends on the level of another predictor. 
Including interaction terms in your model allows you to explore these combined effects and understand more complex relationships between variables. 
In medical research, interactions can be critical, as the impact of a treatment or risk factor might differ across subgroups (for example, by age or sex).

Below is an example of an OLS model fitted using the **rms** package that includes an interaction between Age and Exer. 
This model tests whether the effect of Age on the outcome differs by levels of exercise

```{r}
# Using the built-in dataset from the MASS package
data("survey", package = "MASS")

# Fit an OLS model using the rms package with an interaction between Age and Exer.
fit_interact <- ols(Wr.Hnd ~ Age * Exer + Sex, data = survey)

# Generate a model summary that includes the interaction term
summary_interact <- modelsummary_rms(fit_interact)
knitr::kable(summary_interact)
```

# Models with Interactions and RCS

```{r}
# Using the built-in dataset from the MASS package
data("survey", package = "MASS")

# Fit an OLS model with a restricted cubic spline for Age and an interaction between Age and Exer.
fit_spline_interact <- ols(Wr.Hnd ~ rcs(Age, 4) * Exer + Sex, data = survey)

# Generate a model summary with default RCS output
summary_spline_interact <- modelsummary_rms(fit_spline_interact,
                                            combine_ci = FALSE,
                                            hide_rcs_coef = TRUE,
                                            rcs_overallp = TRUE
                                            )

# Format the output as a nice table
knitr::kable(summary_spline_interact)
```


